---
title: "Customer Lifetime Value in R: A Third of Your Best Customers Have Already Left"
author: Mike Nguyen
date: '2026-09-12'
slug: customer-lifetime-value-in-r
categories:
  - R
  - Python
tags:
  - customer lifetime value
  - CLV
  - Pareto/NBD
  - CLVTools
  - marketing analytics
  - retention
subtitle: 'Rank your customers by what they have spent, and the median customer in the top decile who the model calls inactive goes on to spend nothing.'
summary: 'On the CDNOW panel, with a real 39 week holdout as ground truth, a Pareto/NBD model flags 32 percent of the top decile by historical spend as probably inactive. The median flagged customer then buys nothing at all, against a median of 130 for the rest of the decile. Extrapolating repeat spending overstates the holdout by 35 percent and the model undershoots it by 16, so the level is not where the argument is strongest. Fitted in R with CLVTools, with a bootstrap confidence interval computed in Python.'
authors: []
lastmod: '2026-09-12T11:00:00-07:00'
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```

Sort your customers by what they have spent and take the top decile. That list is the one
retention budgets get spent against. On real transaction data with a real holdout period, a
third of that list had almost certainly stopped buying before the list was drawn, and the
median customer in that third went on to spend nothing at all.

The problem is that in a non-contractual business nobody cancels. There is no churn event to
observe, so a customer who has quietly left looks exactly like a customer who is between
purchases. Historical spend cannot tell them apart, because it throws away the one thing that
distinguishes them, which is when the spending happened.

```{r data}
library(CLVTools)
data("cdnow")

SPLIT <- as.Date("1997-10-01")   # 39 weeks of history, 39 weeks of holdout
d <- as.data.frame(cdnow)
d$Date <- as.Date(d$Date)
cal <- subset(d, Date <= SPLIT)

c(customers = length(unique(d$Id)),
  transactions = nrow(d),
  weeks = round(as.numeric(max(d$Date) - min(d$Date)) / 7))
```

The CDNOW panel is the standard public test bed for this: every customer made a first
purchase in the opening quarter of 1997, and the panel runs to mid 1998. Splitting it in half
gives 39 weeks to learn from and 39 weeks of real behaviour to be judged against. Nothing
below is simulated.

## What extrapolation does

The naive forecast writes itself when the two windows are the same length. Whatever a
customer spent in the last 39 weeks, assume they spend again in the next 39.

```{r naive}
clv <- clvdata(cdnow, date.format = "ymd", time.unit = "week", estimation.split = 39,
               name.id = "Id", name.date = "Date", name.price = "Price")
fit <- pnbd(clv)
p   <- as.data.frame(predict(fit, verbose = FALSE))
p$hist_spend <- as.numeric(tapply(cal$Price, cal$Id, sum)[as.character(p$Id)])

data.frame(
  method = c("actual holdout", "extrapolate last period", "Pareto/NBD"),
  total  = round(c(sum(p$actual.period.spending), sum(p$hist_spend),
                   sum(p$predicted.period.spending))),
  error  = c("", sprintf("%+.0f%%", 100 * (sum(p$hist_spend) /
                                           sum(p$actual.period.spending) - 1)),
             sprintf("%+.0f%%", 100 * (sum(p$predicted.period.spending) /
                                       sum(p$actual.period.spending) - 1))))
```

```{r lvl, include=FALSE}
naive_err <- 100 * (sum(p$hist_spend) / sum(p$actual.period.spending) - 1)
mod_err   <- 100 * (sum(p$predicted.period.spending) / sum(p$actual.period.spending) - 1)
```

Extrapolation overshoots by `r sprintf("%.0f%%", naive_err)`. The model undershoots by
`r sprintf("%.0f%%", abs(mod_err))`, so it is closer but it is not free of error either, and
that is worth saying out loud before the part where it wins.

That `r sprintf("%.0f%%", naive_err)` also flatters the argument, so here is the correction
against my own headline. Every CDNOW customer makes a first purchase in the opening quarter,
which means the calibration window contains an acquisition purchase that the holdout window
structurally cannot. Strip each customer's first transaction and extrapolate only repeat
spending, and the comparison becomes fair.

```{r fair}
first_amt <- sapply(split(cal, cal$Id), function(z) z$Price[which.min(z$Date)])
p$repeat_spend <- p$hist_spend - as.numeric(first_amt[as.character(p$Id)])
fair_err <- 100 * (sum(p$repeat_spend) / sum(p$actual.period.spending) - 1)

c(repeat_only_forecast = round(sum(p$repeat_spend)),
  actual = round(sum(p$actual.period.spending)),
  overshoot_pct = round(fair_err))
```

Still an overshoot, still in the same direction, but `r sprintf("%.0f%%", fair_err)` rather
than `r sprintf("%.0f%%", naive_err)`. Use that number as the honest one. The level is not
where this argument is strongest anyway.

## The list that matters

The level is the less interesting failure. The ranking is where money actually gets
allocated. `pnbd` returns `PAlive`, the probability a customer is still active given how much
and how recently they bought.

```{r decile}
top  <- p[order(p$hist_spend, decreasing = TRUE)[1:round(0.10 * nrow(p))], ]
dead <- top$PAlive < 0.5

data.frame(
  group = c("flagged inactive", "still active"),
  customers = c(sum(dead), sum(!dead)),
  mean_holdout   = round(c(mean(top$actual.period.spending[dead]),
                           mean(top$actual.period.spending[!dead])), 2),
  median_holdout = round(c(median(top$actual.period.spending[dead]),
                           median(top$actual.period.spending[!dead])), 2),
  pct_buying_nothing = round(100 * c(mean(top$actual.period.spending[dead] == 0),
                                     mean(top$actual.period.spending[!dead] == 0))))
```

```{r deciletxt, include=FALSE}
pct_dead  <- 100 * mean(dead)
med_dead  <- median(top$actual.period.spending[dead])
med_live  <- median(top$actual.period.spending[!dead])
ratio     <- mean(top$actual.period.spending[!dead]) / mean(top$actual.period.spending[dead])
rev_share <- 100 * sum(top$actual.period.spending[dead]) / sum(top$actual.period.spending)
```

`r sprintf("%.0f%%", pct_dead)` of the top decile is flagged. Those customers go on to earn
`r sprintf("%.0f%%", rev_share)` of the decile's holdout revenue, they spend
`r sprintf("%.1f", ratio)` times less on average, and the median one spends
`r sprintf("%.0f", med_dead)` dollars against `r sprintf("%.0f", med_live)` for the rest. The
median is the number to look at, because the mean is held up by a handful of flagged customers
who did come back. Most of them simply do not.

```{r figure, echo=FALSE, fig.cap="Within the top decile by historical spend, what customers actually went on to spend, binned by the model's probability that they were still active. The mean is noisy in the small low-probability bins because a few flagged customers did come back. The median does not move."}
library(ggplot2)
# pmin() because pnbd returns PAlive as 1 + 3.6e-15 for one customer here, and
# cut() sends anything above the top break to NA. Indexing with a logical vector
# then pushes that NA into every bin and every statistic comes back NA.
top$bin <- cut(pmin(top$PAlive, 1), breaks = seq(0, 1, by = 0.2), include.lowest = TRUE,
               labels = c("0.0-0.2", "0.2-0.4", "0.4-0.6", "0.6-0.8", "0.8-1.0"))
bin_stat <- do.call(rbind, lapply(split(top$actual.period.spending, top$bin), function(z) {
  data.frame(n = length(z), mean = mean(z), median = median(z))
}))
bin_stat$bin <- rownames(bin_stat)
long <- rbind(data.frame(bin = bin_stat$bin, n = bin_stat$n,
                         stat = "mean",   value = bin_stat$mean),
              data.frame(bin = bin_stat$bin, n = bin_stat$n,
                         stat = "median", value = bin_stat$median))

ggplot(long, aes(bin, value, fill = stat)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66) +
  geom_text(data = bin_stat, aes(bin, mean, label = paste0("n = ", n)),
            inherit.aes = FALSE, vjust = -0.7, size = 3.1, colour = "grey35") +
  scale_fill_manual(values = c(mean = "#8fb8cd", median = "#1f5f84")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.16))) +
  labs(x = "model probability the customer is still active",
       y = "actual holdout spend", fill = NULL,
       title = "Same top-decile list, very different futures",
       subtitle = "CDNOW, 39 week holdout, actual spending rather than predicted") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(), legend.position = "bottom")
```

## Why recency carries the signal

This is not the model asserting something. Hold purchase count fixed and let only recency
vary, and the difference is already in the raw data.

```{r recency}
p$cal_n <- as.numeric(tapply(cal$Price, cal$Id, length)[as.character(p$Id)])
lastbuy <- tapply(cal$Date, cal$Id, max)
p$recency_wk <- as.numeric(SPLIT - as.Date(lastbuy[as.character(p$Id)],
                                           origin = "1970-01-01")) / 7

do.call(rbind, lapply(2:4, function(k) {
  s <- subset(p, cal_n == k)
  recent <- s$recency_wk <= median(s$recency_wk)
  data.frame(purchases = k, customers = nrow(s),
             recent_half = round(mean(s$actual.period.spending[recent]), 2),
             stale_half  = round(mean(s$actual.period.spending[!recent]), 2),
             ratio = round(mean(s$actual.period.spending[recent]) /
                           mean(s$actual.period.spending[!recent]), 2))
}))
```

Two customers, both with four purchases on the books, and the one who bought more recently is
worth multiples of the one who did not. Historical spend scores them identically. That is the
whole gap, and it is why the fix is a model that uses recency rather than a better
spreadsheet.

A bootstrap on the headline gap, computed in Python so the resampling is independent of the R
code above. Customers are resampled, model predictions are held fixed, so this is uncertainty
in the population difference and not in the fitted parameters.

```{r bootprep, include=FALSE}
top_spend <- top$actual.period.spending
top_flag  <- as.integer(dead)
```

```{python boot}
import numpy as np

spend = np.array(r.top_spend, dtype=float)
flag  = np.array(r.top_flag, dtype=bool)
rng   = np.random.default_rng(20260912)

obs = spend[~flag].mean() / spend[flag].mean()
idx = rng.integers(0, len(spend), size=(4000, len(spend)))
reps = []
for row in idx:
    s, f = spend[row], flag[row]
    if f.sum() > 0 and (~f).sum() > 0:
        reps.append(s[~f].mean() / s[f].mean())
lo, hi = np.percentile(reps, [2.5, 97.5])
print(f"active / inactive mean spend ratio: {obs:.2f}")
print(f"95% bootstrap interval: {lo:.2f} to {hi:.2f}  ({len(reps)} resamples)")
```

## What this does not fix

The model is not a forecast of the total, and the table above says so: it came in
`r sprintf("%.0f%%", abs(mod_err))` under the real holdout figure. Pareto/NBD is built to rank
customers and to estimate who is still there, and that is what it is good at.

The ranking gain is also narrower than the headline suggests. Targeting the top decile by
model prediction rather than by historical spend captures only a couple of extra points of
holdout revenue, because the two lists overlap heavily. The value is not a different list, it
is knowing which names on the same list to stop spending on.

```{r capture}
capture <- function(score, frac) {
  sel <- order(score, decreasing = TRUE)[1:ceiling(frac * nrow(p))]
  100 * sum(p$actual.period.spending[sel]) / sum(p$actual.period.spending)
}
data.frame(targeted = c("top 5%", "top 10%", "top 20%"),
           by_historical_spend = round(sapply(c(.05, .10, .20),
                                              function(f) capture(p$hist_spend, f)), 1),
           by_model = round(sapply(c(.05, .10, .20),
                                   function(f) capture(p$predicted.period.spending, f)), 1))
```

And none of this applies to a contractual business. If customers sign up and cancel, you
observe churn directly, a survival model is the right tool, and the whole latent-attrition
apparatus is unnecessary. This is a fix for retail, ecommerce, and anywhere else that people
leave without telling you.

One dataset, one product category, one period. CDNOW is the standard benchmark precisely
because it is public and well studied, not because 1997 compact disc buyers resemble your
customers.

## What to do on Monday

1. Stop ranking customers on historical spend alone. It scores a departed customer and a loyal
   one identically, and the departed one is cheaper to identify than to win back.
2. Split your history and score yourself against a real holdout before trusting any lifetime
   value number. The calculation above took one split and one model fit.
3. Report the probability a customer is still active next to their value, not instead of it.
   The decision is different for a high-value customer at 0.2 than at 0.9.
4. Check whether your business is contractual first. If you can see cancellations, none of
   this is your problem.

The latent attrition models behind this, and where they break, are covered in
[Marketing Research](https://bookdown.org/mike/marketing_research/).
