---
title: "Johnson-Neyman for meta analysis"
author: "Mike Nguyen"
date: '2022-03-03'
slug: johnson-neyman-for-meta-analysis
categories:
- R
- research
tags:
- Academic
- R
subtitle: ''
summary: ''
authors: []
lastmod: '2022-03-03T20:21:15-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r, include = FALSE}
knitr::opts_chunk$set(message = F, warning = F)
```


In applied research, we typically look at moderation effect of a moderator on the relationship between a predictor and an outcome variable. However, looking at the interaction term alone is very misleading. Hence, researcher typically employ the Johnson-Neyman technique in probing the moderation effect. This technique helps find the region where the predictor effect is significant. 

Even though `interactions` package in R is very versatile that it can work with pretty much every model for probing the interaction effect, it couldn't play well with `metafor` package ([for more details on moderation](https://bookdown.org/mike/data_analysis/moderation.html)). Hence, we have to write a custom function for this purpose.

A little reminder of the moderation effect

$$
Y = b_1 X+ b_2 W + b_3 XW
$$

Johnson-Neyman finds the values of the moderators $W$ such that the p-value of the conditional effect of the predictor is equal to the level of significance $\alpha$

$$
t_{crit} = \frac{b_1 + b_3 W}{\sqrt{s^2_{b1} + 2 Ws_{b_1b_3} + W^2 s^2_{b_3}}}
$$

The solution for $W$ is

$$
W = \frac{-2(t_{crit}^2 s_{b_1b_3} - b_1b_3)\pm \sqrt{(2t^2_{crit}s_{b_1b_3}- 2b_1 b_3)^2-4(t_{crit}^2s^2_{b_3} - b_3^2)(t_{crit}^2 s^2_{b_1}-b_1^2)}}{2(t^2_{crit}s^2_{b_3}-b^2_3)}
$$

To implement the custom function, I model the code after [Marko Bachl](https://rpubs.com/bachl/jn-plot)

To get the Johnson-Neyman signficance area 

```{r}
jnt = function(.rma.mv, predictor, moderator, alpha = .05) {
    require(stringi)
    library(tidyverse)
    
    # get coefficient for the predictor
    b1 = .rma.mv$b[predictor, 1]
    # get coefficient for the interaction term
    b3 = .rma.mv$b[paste0(predictor, ":", moderator), 1]
    
    # standard error for the predictor term
    se_b1 = .rma.mv$se[which(rownames(summary(.rma.mv)$b) == predictor)]
    
    # standard error for the interaction term
    se_b3 = .rma.mv$se[which(rownames(summary(.rma.mv)$b) == paste0(predictor, ":", moderator))]
    
    # covariance for the predictor and the interaction term
    COV_b1b3 = .rma.mv$vb[which(rownames(summary(.rma.mv)$b) == predictor), which(rownames(summary(.rma.mv)$b) == paste0(predictor, ":", moderator))]
    
    # the t-critical value
    t_crit = qt(1 - alpha / 2, df.residual(.rma.mv))
    
    
    # see Bauer & Curran, 2005
    a = t_crit ^ 2 * se_b3 ^ 2 - b3 ^ 2
    
    b = 2 * (t_crit ^ 2 * COV_b1b3 - b1 * b3)
    
    c = t_crit ^ 2 * se_b1 ^ 2 - b1 ^ 2
    
    jn = c((-b - sqrt(b ^ 2 - 4 * a * c)) / (2 * a),
           (-b + sqrt(b ^ 2 - 4 * a * c)) / (2 * a))
    
    JN = sort(unname(jn))
    
    # see whether it's inside the two numbers or outside
    
    if (length(JN) != 0){
        # randomly pick a number between the two cutoffs
        test_num <- runif(1, JN[1], JN[2])
        test_theta = b1 + test_num * b3
        test_se_theta = sqrt(se_b1 ^ 2 + 2 * test_num * COV_b1b3 + test_num ^ 2 * se_b3 ^ 2)
        
        test.ci.lo_theta = test_theta - qt(1 - alpha / 2, df.residual(.rma.mv)) * test_se_theta
        test.ci.hi_theta = test_theta + qt(1 - alpha / 2, df.residual(.rma.mv)) * test_se_theta
        if (0 %in% c(test.ci.lo_theta:test.ci.hi_theta)){
            print("inside the range")
        } else {
            print("outside the range")
        }
    }
    
    
    return(JN)
    rm(.rma.mv)
}
```

If the results are 0, 1, 2, then the whole range of the predictors are significant.

```{r}
theta_plot = function(.rma.mv,
                      predictor,
                      moderator,
                      alpha = .05,
                      jn = F) {
    require(dplyr)
    require(ggplot2)
    theme_set(theme_minimal())
    require(stringi)
    
    .data = tibble(
        # get coefficient for the predictor
        b1 = .rma.mv$b[predictor, 1],
        # get coefficient for the interaction term
        b3 = .rma.mv$b[paste0(predictor, ":", moderator), 1],
        
        Z = quantile(.rma.mv$X[, moderator], seq(0, 1, .01)),
        
        theta = b1 + Z * b3,
        
        # standard error for the predictor term
        se_b1 = .rma.mv$se[which(rownames(summary(.rma.mv)$b) == predictor)],
        
        # standard error for the interaction term
        se_b3 = .rma.mv$se[which(rownames(summary(.rma.mv)$b) == paste0(predictor, ":", moderator))],
        
        # covariance for the predictor and the interaction term
        COV_b1b3 = .rma.mv$vb[which(rownames(summary(.rma.mv)$b) == predictor), which(rownames(summary(.rma.mv)$b) == paste0(predictor, ":", moderator))], 
        
        se_theta = sqrt(se_b1 ^ 2 + 2 * Z * COV_b1b3 + Z ^ 2 * se_b3 ^ 2),
        
        ci.lo_theta = theta - qt(1 - alpha / 2, df.residual(.rma.mv)) * se_theta,
        ci.hi_theta = theta + qt(1 - alpha / 2, df.residual(.rma.mv)) * se_theta
    )
    
    
    if (jn) {
        JN = jnt(
            .rma.mv = .rma.mv,
            predictor = predictor,
            moderator = moderator,
            alpha = alpha
        )
        JN_lines = geom_vline(xintercept = JN, linetype = 2)
        JN_regions = ifelse(length(JN) == 0,
                            "no significance regions",
                            paste(round(JN, 2), collapse = "; "))
        Xlab = paste0(moderator, " (JN Significance Regions: ", JN_regions, ")")
    } else {
        Xlab = moderator
        JN_lines = NULL
    }
    .data %>%
        ggplot(aes(x = Z, y = theta)) +
        geom_ribbon(aes(ymin = ci.lo_theta, ymax = ci.hi_theta), fill = "grey70") +
        geom_line() +
        ggtitle(paste(
            "Conditional Effect of",
            predictor,
            "as function of",
            moderator
        )) +
        geom_hline(yintercept = 0, linetype = 2) +
        labs(x = Xlab, y = expression(theta)) +
        JN_lines +
        xlim(min(.rma.mv$X[, moderator]), max(.rma.mv$X[, moderator]))
}
```

Let's try it on a model 

Note: both the moderator and the predictor must be **continuous**

```{r}
library(metafor)
mod <- rma.mv(yi, vi, mods = ~ year * pubstatus, random = ~ 1 | study/esid, data= dat.assink2016)
summary(mod)

jnt(mod, predictor = "year", moderator = "pubstatus")

theta_plot(mod, predictor = "year", moderator = "pubstatus", jn = T)
```
Then we interpret that the impact of year on the outcome variable is only significant when the publication status is greater than 0.95 or less than -0.02 (but this is not possible because the lower limit is 0). 