---
title: "Synthetic Respondents in Conjoint Analysis: The Average Survives, the Price Does Not"
author: Mike Nguyen
date: '2026-09-08'
slug: synthetic-respondents-conjoint-analysis
categories:
  - R
  - Python
tags:
  - conjoint analysis
  - synthetic respondents
  - LLM
  - pricing
  - mixed logit
  - marketing research
subtitle: 'The panel passed every validation check I ran, then priced the product a quarter too low.'
summary: 'LLM synthetic respondents are known to produce response distributions that are too narrow. Nobody writes down what that costs you. Running a conjoint through a mixed logit and into a market simulator, a panel with correct mean price sensitivity and compressed heterogeneity reproduced the human holdout choice shares almost exactly, then set the revenue-maximizing price 26% too low and sized the premium segment at less than half its true value. Simulated in R with logitr, cross-checked in Python by Gauss-Hermite quadrature.'
authors: []
lastmod: '2026-09-08T19:30:00-07:00'
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```

If you buy a panel of LLM synthetic respondents for a conjoint analysis, the validation
report you get back will compare average part-worths and predicted choice shares. Those are
exactly the numbers that survive the failure mode the method actually has, and the price you
set afterwards is not.

The known problem is that LLM respondents produce response distributions that are too narrow.
Maier et al. built their semantic similarity rating method around it, because asking a model
directly for a numeric rating "produces unrealistic response distributions" ([arXiv
2510.08338](https://arxiv.org/abs/2510.08338), October 2025). What nobody writes down is what
a narrow distribution costs you once the study reaches a pricing decision. So I simulated a
conjoint where I picked the true answer myself, compressed the heterogeneity the way the
literature says these panels do, and ran both panels through the same workflow.

```{r truth}
# One population, two panels. Identical MEAN price sensitivity by construction,
# and heterogeneity shrunk by a factor lambda in the synthetic one.
k_mean <- 0.05   # E|price coefficient|, held fixed across every panel below
s_full <- 0.90   # true dispersion in log price sensitivity
mu_f   <- 1.20   # mean part-worth for the premium feature
sg_f   <- 0.90
mu_b   <- 0.50   # mean brand part-worth
sg_b   <- 0.60
asc_none <- -1.5 # utility of not buying
mc     <- 12     # marginal cost, so there is a real price to optimise
LAM    <- 0.40   # the synthetic panel keeps 40% of the true spread

# Price sensitivity is lognormal, which is standard because it keeps the sign
# right. Holding E|b_price| fixed while shrinking the spread requires moving the
# log-mean, so m is a function of s rather than a constant.
log_m <- function(s) log(k_mean) - s^2 / 2
```

The premium feature is worth `r sprintf("$%.0f", mu_f / k_mean)` to the average buyer. Prices
tested in the survey run from \$25 to \$85, three products plus a no-buy option, 300
respondents, 12 tasks each.

```{r simulate}
n_resp <- 300
n_task <- 12
n_prod <- 3
prices <- c(25, 45, 65, 85)

draw_pop <- function(n, lambda, seed) {
  set.seed(seed)
  s <- s_full * lambda
  data.frame(id = seq_len(n),
             bp = -exp(log_m(s) + s * rnorm(n)),
             bf = mu_f + sg_f * lambda * rnorm(n),
             bb = mu_b + sg_b * lambda * rnorm(n))
}

make_design <- function(seed) {
  set.seed(seed)
  n <- n_resp * n_task * n_prod
  d <- data.frame(
    id     = rep(seq_len(n_resp), each = n_task * n_prod),
    task   = rep(rep(seq_len(n_task), each = n_prod), n_resp),
    price  = sample(prices, n, TRUE),
    feat   = rbinom(n, 1, 0.5),
    brandB = rbinom(n, 1, 0.5))
  d$obs <- (d$id - 1) * n_task + d$task
  d
}

sim_choice <- function(design, pop, seed) {
  set.seed(seed)
  b <- pop[match(design$id, pop$id), ]
  d <- design
  d$V <- b$bp * d$price + b$bf * d$feat + b$bb * d$brandB
  none <- unique(d[, c("id", "task", "obs")])
  none$price  <- 0
  none$feat   <- 0
  none$brandB <- 0
  none$V      <- asc_none
  d <- rbind(d[, names(none)], none)
  d <- d[order(d$obs), ]
  d$none     <- as.integer(d$price == 0 & d$feat == 0 & d$brandB == 0)
  d$negPrice <- -d$price
  d$choice   <- 0L
  for (k in split(seq_len(nrow(d)), d$obs)) {
    p <- exp(d$V[k])
    d$choice[k[sample.int(length(k), 1, prob = p / sum(p))]] <- 1L
  }
  d
}

design <- make_design(11)
human  <- sim_choice(design, draw_pop(n_resp, 1.0, 21), 31)
synth  <- sim_choice(design, draw_pop(n_resp, LAM, 21), 31)
```

## The validation that passes

Fit the same mixed logit to each panel, holding out the last two tasks. This is the model a
conjoint study actually reports.

```{r fit, results='hide'}
library(logitr)

fit_panel <- function(d) {
  logitr(data = d[d$task <= 10, ], outcome = "choice", obsID = "obs", panelID = "id",
         pars     = c("negPrice", "feat", "brandB", "none"),
         randPars = c(negPrice = "ln", feat = "n", brandB = "n"),
         numDraws = 200)
}
f_human <- fit_panel(human)
f_synth <- fit_panel(synth)
```

```{r validation}
ch <- coef(f_human)
cs <- coef(f_synth)

# E|b_price| = exp(m + s^2/2) is the interpretable quantity, not the log-mean.
mean_bp <- function(cf) exp(cf["negPrice"] + abs(cf["sd_negPrice"])^2 / 2)

# The check a validation report runs: predict the HUMAN holdout choice shares
# from each fitted model and compare against what the humans actually did.
hold <- human[human$task > 10, ]
pred_share <- function(fit) {
  p <- predict(fit, newdata = hold, obsID = "obs", returnData = FALSE)
  mean(p$predicted_prob[hold$none == 0])
}

data.frame(
  quantity = c("mean price sensitivity", "feature part-worth", "brand part-worth",
               "predicted product share, human holdout"),
  human = round(c(mean_bp(ch), ch["feat"], ch["brandB"], pred_share(f_human)), 4),
  synthetic = round(c(mean_bp(cs), cs["feat"], cs["brandB"], pred_share(f_synth)), 4),
  row.names = NULL)
```

```{r actual}
c(actual_human_holdout_share = round(mean(hold$choice[hold$none == 0]), 4))
```

Every central quantity agrees. The synthetic panel predicts the human holdout share to within
a fraction of a point, and on that particular number it lands closer to the truth than the
human panel's own model does. On this evidence you would sign off on the panel.

Now look at what the same two fits say about spread.

```{r spread}
data.frame(
  parameter = c("log price sensitivity", "feature part-worth", "brand part-worth"),
  human_sd    = round(abs(ch[c("sd_negPrice", "sd_feat", "sd_brandB")]), 3),
  synth_sd    = round(abs(cs[c("sd_negPrice", "sd_feat", "sd_brandB")]), 3),
  row.names = NULL)
```

Price and feature spread are roughly a third of the truth, and brand spread is gone entirely.
On brand the panel is one customer, repeated 300 times.

## The decision that breaks

Conjoint output is not the deliverable. The market simulator is. Take each fitted model,
put the premium product up against two fixed competitors, and find the price that maximises
contribution.

```{r simulator}
comp <- data.frame(price = c(35, 55), feat = c(0, 1), brandB = c(1, 1))
grid <- seq(15, 200, by = 0.5)

simulate_market <- function(cf, ndraw = 8e4, seed = 7) {
  set.seed(seed)
  z  <- matrix(rnorm(3 * ndraw), ncol = 3)
  bp <- -exp(cf["negPrice"] + abs(cf["sd_negPrice"]) * z[, 1])
  bf <- cf["feat"]   + abs(cf["sd_feat"])   * z[, 2]
  bb <- cf["brandB"] + abs(cf["sd_brandB"]) * z[, 3]
  e1 <- exp(bp * comp$price[1] + bf * comp$feat[1] + bb * comp$brandB[1])
  e2 <- exp(bp * comp$price[2] + bf * comp$feat[2] + bb * comp$brandB[2])
  d0 <- e1 + e2 + exp(cf["none"])
  share <- vapply(grid, function(p) {
    ef <- exp(bp * p + bf)
    mean(ef / (ef + d0))
  }, 0)
  list(share = share,
       opt   = grid[which.max(share * (grid - mc))],
       seg   = 100 * mean((bf / -bp) > 50))
}

truth_cf <- c(negPrice = log_m(s_full), feat = mu_f, brandB = mu_b, none = asc_none,
              sd_negPrice = s_full, sd_feat = sg_f, sd_brandB = sg_b)

m_truth <- simulate_market(truth_cf)
m_human <- simulate_market(ch)
m_synth <- simulate_market(cs)

data.frame(
  source              = c("truth", "human panel", "synthetic panel"),
  optimal_price       = c(m_truth$opt, m_human$opt, m_synth$opt),
  premium_segment_pct = round(c(m_truth$seg, m_human$seg, m_synth$seg), 1),
  row.names = NULL)
```

```{r headline, include=FALSE}
err_price <- 100 * (m_truth$opt - m_synth$opt) / m_truth$opt
err_vs_h  <- 100 * (m_human$opt - m_synth$opt) / m_human$opt
seg_ratio <- m_truth$seg / m_synth$seg
i_t <- which.max(m_truth$share * (grid - mc))
i_s <- which.max(m_synth$share * (grid - mc))
kept <- 100 * (m_truth$share[i_s] * (grid[i_s] - mc)) /
               (m_truth$share[i_t] * (grid[i_t] - mc))
```

The synthetic panel sets the price `r sprintf("%.0f%%", err_price)` below the truth and
`r sprintf("%.0f%%", err_vs_h)` below what the human panel recommends, and it sizes the
premium segment at `r sprintf("%.0f%%", 100 / seg_ratio)` of its true value. Contribution at the
synthetic price is `r sprintf("%.1f%%", kept)` of the best available, so the money lost is
real but modest. That flatness is the trap. Profit curves are flat near their peak, which is
precisely why a price recommendation can be a quarter too low without anything downstream
looking obviously broken.

## Why the average survives

Compressing heterogeneity leaves the mean where it was and removes the tails. Demand at a
high price is made entirely of the price-insensitive tail, so it is the first thing to go.

```{r decay}
k <- match(c(45, 65, 85, 125), grid)
data.frame(price = grid[k],
           truth_share = round(m_truth$share[k], 4),
           synth_share = round(m_synth$share[k], 4),
           ratio       = round(m_synth$share[k] / m_truth$share[k], 2))
```

At the prices the survey tested the two panels agree. Past them they diverge, and a pricing
study is an extrapolation past the tested range by construction. Sweeping the compression
factor shows how cleanly the quantities separate.

```{r sweep, fig.cap="Mean price sensitivity is flat by construction. Everything that depends on the spread is not."}
sweep_lambda <- function(lambda, ndraw = 8e4, seed = 7) {
  set.seed(seed)
  z  <- matrix(rnorm(3 * ndraw), ncol = 3)
  s  <- s_full * lambda
  bp <- -exp(log_m(s) + s * z[, 1])
  bf <- mu_f + sg_f * lambda * z[, 2]
  bb <- mu_b + sg_b * lambda * z[, 3]
  e1 <- exp(bp * comp$price[1] + bf * comp$feat[1] + bb * comp$brandB[1])
  e2 <- exp(bp * comp$price[2] + bf * comp$feat[2] + bb * comp$brandB[2])
  d0 <- e1 + e2 + exp(asc_none)
  share <- vapply(grid, function(p) {
    ef <- exp(bp * p + bf)
    mean(ef / (ef + d0))
  }, 0)
  data.frame(lambda = lambda,
             mean_price_coef = mean(-bp),
             opt_price = grid[which.max(share * (grid - mc))],
             premium_seg = 100 * mean((bf / -bp) > 50))
}

lams <- seq(1, 0.2, by = -0.1)
sw   <- do.call(rbind, lapply(lams, sweep_lambda))
base <- sw[sw$lambda == 1, ]
round(sw, 4)
```

```{r sweepplot, echo=FALSE, fig.cap="Only the validated quantity is stable. Optimal price and segment size fall away as the panel loses spread."}
library(ggplot2)
plot_df <- rbind(
  data.frame(lambda = sw$lambda, pct = 100 * sw$mean_price_coef / base$mean_price_coef,
             q = "mean price sensitivity"),
  data.frame(lambda = sw$lambda, pct = 100 * sw$opt_price / base$opt_price,
             q = "revenue-maximising price"),
  data.frame(lambda = sw$lambda, pct = 100 * sw$premium_seg / base$premium_seg,
             q = "premium segment size"))
plot_df$q <- factor(plot_df$q, levels = c("mean price sensitivity",
                                          "revenue-maximising price",
                                          "premium segment size"))
ggplot(plot_df, aes(lambda, pct, colour = q)) +
  geom_hline(yintercept = 100, linetype = 3, colour = "grey55") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.8) +
  scale_x_reverse(breaks = lams) +
  scale_y_continuous(limits = c(0, 115)) +
  scale_colour_manual(values = c("#2c7fb8", "#d95f0e", "#c51b8a")) +
  labs(x = "share of true heterogeneity retained by the panel",
       y = "percent of the true value", colour = NULL,
       title = "What a narrow synthetic panel keeps, and what it loses") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The blue line is what a validation report checks. The other two are what the study is for.

An independent check, because a Monte Carlo integral deserves one. The same optimum computed
in Python by three-dimensional Gauss-Hermite quadrature, a deterministic method with no
sampling in it at all.

```{python quadrature}
import numpy as np

k_mean, s_full = r.k_mean, r.s_full
mu_f, sg_f, mu_b, sg_b = r.mu_f, r.sg_f, r.mu_b, r.sg_b
asc_none, mc = r.asc_none, r.mc
cp, cf_, cb = np.array([35., 55.]), np.array([0., 1.]), np.array([1., 1.])

def opt_price(lam, n=24):
    x, w = np.polynomial.hermite_e.hermegauss(n)
    w = w / w.sum()
    s = s_full * lam
    bp = -np.exp(np.log(k_mean) - s**2 / 2 + s * x)
    bf = mu_f + sg_f * lam * x
    bb = mu_b + sg_b * lam * x
    BP, BF, BB = bp[:, None, None], bf[None, :, None], bb[None, None, :]
    W = w[:, None, None] * w[None, :, None] * w[None, None, :]
    d0 = (np.exp(BP * cp[0] + BF * cf_[0] + BB * cb[0])
          + np.exp(BP * cp[1] + BF * cf_[1] + BB * cb[1]) + np.exp(asc_none))
    grid = np.arange(15, 200.5, 0.5)
    share = np.array([(W * (np.exp(BP * p + BF) / (np.exp(BP * p + BF) + d0))).sum()
                      for p in grid])
    return grid[(share * (grid - mc)).argmax()], (w * (-bp)).sum()

for lam in [1.0, 0.8, 0.6, 0.4, 0.2]:
    p, mb = opt_price(lam)
    print(f"lambda {lam:.1f}  optimal price {p:6.1f}   E|b_price| {mb:.5f}")
```

Quadrature returns the same optimal price as the R simulation at every compression level, and
the mean price coefficient sits on `r sprintf("%.3f", k_mean)` throughout. The collapse is a
property of the model, not of the random draws.

## The diagnostic

The compression is recoverable from choice data, which means you can test for it. Fit the
mixed logit to your synthetic panel and to a small human pilot, then divide the estimated
standard deviations.

```{r diagnostic}
lambda_hat <- abs(cs[c("sd_negPrice", "sd_feat", "sd_brandB")]) /
              abs(ch[c("sd_negPrice", "sd_feat", "sd_brandB")])
round(c(lambda_hat, "true lambda" = LAM), 3)
```

Read this as a screen and not as a correction factor. All three ratios land well below 1, so
the screen fires, which is the job. As point estimates they are poor: only the price ratio is
close to the true `r LAM` (`r sprintf("%.2f", lambda_hat[1])`), the feature ratio overshoots
at `r sprintf("%.2f", lambda_hat[2])`, and the brand ratio is
`r sprintf("%.2f", lambda_hat[3])` because a standard deviation near zero is barely
identified. The parameter that collapses hardest is the one whose collapse is hardest to
measure, so this tells you to keep humans in the study rather than telling you how much to
rescale by.

## What this post did not measure

I did not query a language model. This post takes the narrowness that the literature
documents, applies it as a known compression factor, and prices the consequence. It is a
statement about what a documented failure mode costs, not new evidence that the failure mode
exists. If your panel reproduces human spread, none of this applies to it, and methods
designed to fix exactly this (semantic similarity rating is the current one) exist for that
reason.

Three more limits worth naming. The compression here is uniform across attributes, and a real
panel will lose more spread on some than others. Hierarchical Bayes estimation adds its own
shrinkage on top, so a real workflow is likely to be worse than this rather than better. And
the truth here is a mixed logit, so I have given the model the correct functional form and
still lost the price.

## What to do on Monday

1. Do not validate a synthetic panel on means, rankings, or holdout hit rates alone. Those
   are the statistics that survive the failure.
2. Compare the estimated standard deviations against a human pilot, not just the estimated
   means. The ratio is noisy, so use it as a flag rather than as a rescaling factor.
3. Check where your decision sits relative to the tested attribute range. Interpolation is
   fine. Pricing is extrapolation.
4. If the deliverable is a segment size or a price, keep humans in the study. A small pilot
   is enough to catch the collapse, and it is much cheaper than the price you would have set.

The mixed logit machinery behind all of this, and how to run a conjoint properly, is in
[Marketing Research](https://bookdown.org/mike/marketing_research/).
