---
title: "How to Benchmark R Code Correctly: I Raced a Function Against an Identical Copy of Itself"
author: Mike Nguyen
date: '2026-09-01'
slug: how-to-benchmark-r-code-correctly
categories:
  - R
  - Python
tags:
  - benchmarking
  - performance
  - experimental design
  - microbenchmark
  - reproducibility
subtitle: 'One of them won, and which one depended only on the order I listed them in.'
summary: 'A benchmark is an experiment, and most benchmarks are not randomized. Racing a function against a byte-identical copy of itself, the block ordering that bench::mark uses hands one slot an artifact worth more than most published speedups, and its size is set by the state of the R session rather than by the code, so you cannot subtract it off. Randomizing the execution order removes it in every session tested. Measured in R across fresh sessions and reproduced in Python.'
authors: []
lastmod: '2026-09-01T09:00:00-07:00'
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.0, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```

```{r helpers, include=FALSE}
# Every number in the prose below is an inline expression reading these, so the
# text cannot drift away from what the code actually produced on this machine.
pct  <- function(v) sprintf("%.1f%%", abs(v - 1) * 100)
dirw <- function(v) if (v > 1) "slower" else "faster"
```

A benchmark is an experiment. It has treatments (the implementations you are comparing), units
(the individual timed calls), and a response (elapsed time). Almost nobody treats it
that way, and the cost is that many speed claims you read are measuring the order
the code ran in rather than the code.

Here is the cheapest demonstration. I raced a function against a byte-identical copy
of itself, and swapped which copy got listed first. If a slot rather than a function is winning,
the first slot wins either way.

```{r null-race}
library(microbenchmark)
set.seed(2026)

impl_a <- function(x) {
  y <- x * 2
  z <- cumsum(y)
  z[length(z)]
}
impl_b <- function(x) {
  y <- x * 2
  z <- cumsum(y)
  z[length(z)]
}
stopifnot(identical(body(impl_a), body(impl_b)))

x <- runif(1e5)

# Ratio of the median time of whichever function was LISTED FIRST to the second.
# Anything other than 1 is an artifact, because the two functions are the same code.
race <- function(order, first) {
  mb <- if (first == "a") {
    microbenchmark(a = impl_a(x), b = impl_b(x), times = 40,
                   control = list(order = order))
  } else {
    microbenchmark(b = impl_b(x), a = impl_a(x), times = 40,
                   control = list(order = order))
  }
  m <- tapply(mb$time, mb$expr, median)
  as.numeric(m[[first]] / m[[setdiff(c("a", "b"), first)]])
}

reps <- 40
race_grid <- expand.grid(rep = seq_len(reps), order = c("block", "random"),
                         first = c("a", "b"), stringsAsFactors = FALSE)
race_grid$ratio <- mapply(race, race_grid$order, race_grid$first)
```

```{r null-table, echo=FALSE}
summarise_cell <- function(d, label) {
  tt <- t.test(d$ratio, mu = 1)
  data.frame(
    Ordering = d$order[1],
    `Listed first` = label,
    `Mean ratio` = sprintf("%.4f", mean(d$ratio)),
    `95% CI` = sprintf("[%.4f, %.4f]", tt$conf.int[1], tt$conf.int[2]),
    `p vs 1` = format.pval(tt$p.value, digits = 2),
    `First slot slower` = sprintf("%.0f%%", 100 * mean(d$ratio > 1)),
    check.names = FALSE)
}
rows <- list()
for (o in c("block", "random")) {
  for (f in c("a", "b")) {
    rows[[length(rows) + 1]] <- summarise_cell(
      race_grid[race_grid$order == o & race_grid$first == f, ], f)
  }
  rows[[length(rows) + 1]] <- summarise_cell(race_grid[race_grid$order == o, ],
                                             "either (pooled)")
}
knitr::kable(do.call(rbind, rows), row.names = FALSE)
```

```{r null-values, include=FALSE}
pooled <- function(o) {
  d <- race_grid$ratio[race_grid$order == o]
  tt <- t.test(d, mu = 1)
  list(mean = mean(d), lo = tt$conf.int[1], hi = tt$conf.int[2], p = tt$p.value,
       worse = 100 * mean(d > 1))
}
blk <- pooled("block")
rnd <- pooled("random")
```

Under block ordering the function listed first is measured as `r pct(blk$mean)`
`r dirw(blk$mean)` than an identical copy of itself, and it comes out that way in
`r sprintf("%.0f%%", max(blk$worse, 100 - blk$worse))` of the races. The gap survives the swap, so
it belongs to the slot and not to the function. Under random ordering the same comparison returns
a mean ratio of `r sprintf("%.4f", rnd$mean)`, interval
`r sprintf("[%.4f, %.4f]", rnd$lo, rnd$hi)`, which is the null answer a benchmark of identical
code is supposed to give.

```{r null-plot, echo=FALSE, fig.cap="Two identical functions raced against each other under each ordering. Randomizing the execution order centres the distribution on zero. Block ordering moves it off."}
library(ggplot2)
ggplot(race_grid, aes(x = ratio, fill = order)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(xintercept = 1, linewidth = 0.4) +
  scale_x_continuous(labels = function(v) sprintf("%+.0f%%", (v - 1) * 100)) +
  labs(x = "measured time penalty of the first-listed function against an identical copy",
       y = "density", fill = "execution order") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

## Which ordering is your package using

The two benchmarking packages in common use disagree about this, and neither announces it. You can
just look, by making the expressions log themselves.

```{r order-log}
run_log <- character()
tag_a <- function() {
  run_log <<- c(run_log, "a")
  NULL
}
tag_b <- function() {
  run_log <<- c(run_log, "b")
  NULL
}

run_log <- character()
invisible(bench::mark(a = tag_a(), b = tag_b(), iterations = 6, check = FALSE))
order_bench <- paste(run_log, collapse = " ")

run_log <- character()
invisible(microbenchmark(a = tag_a(), b = tag_b(), times = 6))
order_mb <- paste(run_log, collapse = " ")

cat("bench::mark    :", order_bench, "\n")
cat("microbenchmark :", order_mb, "\n")
```

`bench::mark` calls each expression once up front and then runs every iteration of the first
before it touches the second. That is block ordering, and there is no argument to change it.
`microbenchmark` shuffles, because its `control$order` defaults to `"random"`. That default is the whole
difference between the two halves of the table.

## The size belongs to the session, not to the code

A constant bias would be survivable, because you could measure it once and subtract it off. Here
is the same race run in eight fresh R sessions, each one started clean.

```{r sessions}
library(callr)

one_session <- function(seed) {
  callr::r(function(seed) {
    library(microbenchmark)
    set.seed(seed)
    impl_a <- function(x) {
      y <- x * 2
      z <- cumsum(y)
      z[length(z)]
    }
    impl_b <- function(x) {
      y <- x * 2
      z <- cumsum(y)
      z[length(z)]
    }
    x <- runif(1e5)
    race <- function(order) {
      mb <- microbenchmark(a = impl_a(x), b = impl_b(x), times = 40,
                           control = list(order = order))
      m <- tapply(mb$time, mb$expr, median)
      as.numeric(m[["a"]] / m[["b"]])
    }
    # Interleaved, so both orderings meet the same session conditions.
    out <- vapply(1:16, function(i) c(block = race("block"), random = race("random")),
                  numeric(2))
    rowMeans(out)
  }, args = list(seed = seed))
}

sessions <- as.data.frame(t(vapply(1:8, one_session, numeric(2))))
round(sessions, 4)
```

```{r session-values, include=FALSE}
bs <- sessions$block
rs <- sessions$random
n_same <- sum(sign(bs - 1) == sign(blk$mean - 1))
same_txt <- if (n_same == length(bs)) "all eight" else sprintf("%d of the eight", n_same)
ratio_txt <- sprintf("%.0f", abs(blk$mean - 1) / max(abs(mean(bs) - 1), 1e-6))
```

Same code, same machine, same afternoon. The artifact appears in `r same_txt`, leaning the way this
document did, at `r sprintf("%+.1f%%", (min(bs) - 1) * 100)` to
`r sprintf("%+.1f%%", (max(bs) - 1) * 100)`.

Now compare the session you are reading the output of. The table at the top of this page came out
of the R session that knitted it, and it measured `r pct(blk$mean)` against the `r pct(mean(bs))`
these clean sessions average, a factor of `r ratio_txt`. A long-lived session with a big heap is a
much worse place to benchmark than a fresh one, and a knitted report or a day-old RStudio session
is exactly that. The randomized column stays inside
`r sprintf("%+.1f%%", (min(rs) - 1) * 100)` to `r sprintf("%+.1f%%", (max(rs) - 1) * 100)`
throughout. It does not care.

## What block ordering is aliasing

The first expression is not special. Machine speed drifts over the life of a benchmark, and block
ordering hands the first half of that drift to one expression and the second half to the other.
Time one expression alone and the structure is already there.

```{r drift}
n <- 100
reps_d <- 200
prof <- matrix(NA_real_, reps_d, n)
half <- numeric(reps_d)
half_mean <- numeric(reps_d)
first <- 1:(n / 2)
second <- (n / 2 + 1):n
for (i in seq_len(reps_d)) {
  tt <- as.numeric(microbenchmark(impl_a(x), times = n)$time)
  prof[i, ] <- tt / median(tt)
  half[i]      <- median(tt[first]) / median(tt[second])
  half_mean[i] <- mean(tt[first]) / mean(tt[second])
}
t_half <- t.test(half, mu = 1)
c(median_based = mean(half), mean_based = mean(half_mean),
  ci_low = t_half$conf.int[1], ci_high = t_half$conf.int[2])
```

```{r drift-values, include=FALSE}
avg_prof <- colMeans(prof)
spikes <- which(avg_prof > 1.5)
spike_gap <- if (length(spikes) > 1) round(mean(diff(spikes))) else NA_integer_
spike_peak <- max(avg_prof)
agree <- sign(mean(half) - 1) == sign(blk$mean - 1)
sig_txt <- if (t_half$p.value < 0.05) {
  "and the interval clears 1"
} else {
  "though at this sample size the interval still includes 1"
}
agree_txt <- if (agree) {
  "points the same way as the head to head race, as it should if the race reads position not code"
} else {
  "points against the head to head race here, so drift is one contributor not the whole story"
}
```

One expression, one hundred consecutive calls, no competitor. The first fifty calls
run `r pct(mean(half))` `r dirw(mean(half))` than the second fifty
`r sig_txt` (p = `r format.pval(t_half$p.value, digits = 2)`). That `r agree_txt`.

```{r drift-plot, echo=FALSE, fig.cap=paste0("Mean time by position within a run, normalised to the run median and averaged over ", reps_d, " runs. The regular spikes are garbage collection. Block ordering splits this sequence down the middle and charges each half to a different expression.")}
prof_df <- data.frame(position = seq_len(n), rel = colMeans(prof))
ggplot(prof_df, aes(position, rel)) +
  annotate("rect", xmin = 0.5, xmax = n / 2 + 0.5, ymin = -Inf, ymax = Inf,
           alpha = 0.10, fill = "steelblue") +
  annotate("text", x = n / 4, y = max(prof_df$rel), label = "charged to the first expression",
           vjust = 1.2, size = 3.4, colour = "grey30") +
  annotate("text", x = 3 * n / 4, y = max(prof_df$rel), label = "charged to the second",
           vjust = 1.2, size = 3.4, colour = "grey30") +
  geom_line(linewidth = 0.5, colour = "grey25") +
  geom_hline(yintercept = 1, linetype = 2, linewidth = 0.4) +
  labs(x = "call number within the run", y = "time relative to the run median") +
  theme_minimal(base_size = 12)
```

There are two shapes here. A few calls cost up to `r sprintf("%.1f", spike_peak)` times the median,
spaced about `r spike_gap` calls apart, which is garbage collection on a cycle. Underneath them the
baseline drifts on its own. Medians step over the spikes and means do not, so the two read
different things, and here they agree: `r sprintf("%.4f", mean(half))` against
`r sprintf("%.4f", mean(half_mean))`. The effect survives either summary statistic. Neither shape
is a property of the code, and block ordering hands the first half of both to whichever expression
you listed first.

## Reading a real comparison

Multiplication against addition is a comparison people actually run. The move that makes it
readable is to drop a copy of one candidate into the same benchmark as a negative control.

```{r real-pair}
xx <- runif(1e6)
reps_r <- 40
r_mul <- r_null <- numeric(reps_r)
for (i in seq_len(reps_r)) {
  mb <- microbenchmark(add = xx + xx, mul = 2 * xx, add_copy = xx + xx, times = 40)
  m <- tapply(mb$time, mb$expr, median)
  r_mul[i]  <- m[["mul"]] / m[["add"]]
  r_null[i] <- m[["add_copy"]] / m[["add"]]
}
rbind(`2 * x  vs  x + x` = c(mean = mean(r_mul), t.test(r_mul, mu = 1)$conf.int,
                            worst_single_run = max(abs(r_mul - 1))),
      `x + x  vs  x + x` = c(mean = mean(r_null), t.test(r_null, mu = 1)$conf.int,
                            worst_single_run = max(abs(r_null - 1))))
```

```{r real-values, include=FALSE}
tm <- t.test(r_mul, mu = 1)
effect <- abs(mean(r_mul) - 1)
noise  <- max(abs(r_null - 1))
verdict <- if (noise > effect) {
  "larger than the effect, so one run cannot resolve this comparison"
} else {
  "smaller than the effect, so one run can see this gap, but only because the gap is large"
}
```

`2 * xx` really is faster than `xx + xx`, by `r pct(mean(r_mul))` on average, and across 40
randomized races the interval `r sprintf("[%.3f, %.3f]", tm$conf.int[1], tm$conf.int[2])` clears 1
comfortably. The control row is the useful part. Two identical expressions in the same benchmark
were off by as much as `r sprintf("%.1f%%", noise * 100)` in a single run, which is
`r verdict`. That is your resolution limit for this machine on this day, and it costs one line.

## The same design flaw in Python

This is not an R problem. It is a problem with running all of A and then all of B, which is what
timing tools do by default in every language.

```{r py-setup, include=FALSE}
py_reps <- 60L
```

```{python py-null}
import random
import statistics
import time
import numpy as np

rng = np.random.default_rng(0)
xp = rng.random(100_000)

def impl_a():
    y = xp * 2.0
    return np.cumsum(y)[-1]

def impl_b():
    y = xp * 2.0
    return np.cumsum(y)[-1]

def timed(f):
    t0 = time.perf_counter_ns()
    f()
    return time.perf_counter_ns() - t0

def race(order, n=40, seed=0):
    plan = [("first", impl_a)] * n + [("second", impl_b)] * n
    if order == "random":
        random.Random(seed).shuffle(plan)
    t = {"first": [], "second": []}
    for label, f in plan:
        t[label].append(timed(f))
    return statistics.median(t["first"]) / statistics.median(t["second"])

for _ in range(30):
    _warm = impl_a()

reps = int(r.py_reps)
res = {o: [race(o, seed=i) for i in range(reps)] for o in ("block", "random")}

for o in ("block", "random"):
    v = res[o]
    m = statistics.mean(v)
    se = statistics.stdev(v) / len(v) ** 0.5
    print(f"{o:7s} mean {m:.4f}  95% CI [{m - 1.96 * se:.4f}, {m + 1.96 * se:.4f}]"
          f"  sd {statistics.stdev(v):.4f}")

print(f"variance ratio, block over random = "
      f"{statistics.variance(res['block']) / statistics.variance(res['random']):.2f}")
```

The symptom differs and the diagnosis does not. NumPy does not allocate through R's collector, so
the centre of the Python distribution sits close to 1. What block ordering costs there is
precision. Randomizing shrinks the variance by the factor printed above, because consecutive calls
inside a block are correlated, and a block of 40 correlated calls is worth far fewer than 40
independent ones.

## The checklist

1. **Randomize the execution order.** In R that means `microbenchmark` at its default, or
   `control = list(order = "random")` if you have changed it. `bench::mark` cannot do it, so reach
   for it because of its memory accounting rather than for close races.
2. **Put a negative control in every benchmark.** A second copy of one candidate under a different
   name. Whatever gap it reports against its own twin is your noise floor for that machine on that
   day.
3. **Repeat the whole benchmark, not just the iterations inside it.** Iterations within a run are
   correlated. Runs are much closer to independent, and their spread is what belongs in your
   interval.
4. **Report an interval.** A ratio of two medians with nothing around it is not a result.
5. **Benchmark from a fresh session.** The loaded session that knitted this page carried an
   artifact `r ratio_txt` times larger than clean ones running the same code. `callr::r()` gives
   you a clean session in one line.
6. **Refuse to call anything under the noise floor a win.** Most published single-digit speedups
   would not survive their own negative control.

## What this post did not measure

One machine, one operating system, one R build, one afternoon. The percentages are this machine
talking rather than a constant of nature. What transfers is the structure: block ordering aliases
within-run drift onto the comparison, randomizing removes the alias for free, and a negative
control tells you what your setup can resolve.

The gap between the loaded session and the clean ones is measured but not explained. Heap size is
the obvious suspect, but adding ballast to a clean session did not reproduce it, so something else
is doing the work.

I also did not test whether the numbers `bench::mark` prints inherit this as badly. It uses a
different timer and reports the minimum alongside the median, and a minimum is far more robust to a
spike than a median is. That deserves its own post.

Randomization, blocking, negative controls, and enough replication to see the effect you care
about are not benchmarking ideas. They are experimental design, and they are the subject of
[Experimental Design](https://link.springer.com/book/9783032018380), volume 4 of A Guide on Data
Analysis.
