---
title: reproducible example (reprex)
author: ''
date: '2021-04-03'
slug: reproducible-example-reprex
categories:
  - R
tags:
  - R
  - reproducible
subtitle: ''
summary: ''
authors: []
lastmod: '2021-04-03T13:06:02-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

install once per computer

```{r, eval = FALSE}
install.packages("reprex")
```
or part of `tidyverse

```{r, eval=FALSE}
install.packages("tidyverse")
```


You can either call it in every session 
```{r, eval=FALSE}
library(reprex)
```

or put it in `.Rprofile` for each machine to run whenever you open R

```{r, eval = FALSE}
if (interactive()){
  suppressMessages(require(reprex))
}
options(
  reprex.advertise = FALSE, 
  reprex.si = TRUE, 
  reprex.style = TRUE,
  reprex.comment = "#;-",
  reprex.tidyverse_quiet = FALSE
)
```


to see your `.Rprofile`

```{r, eval = FALSE}
usethis::edit_r_profile()
```

Since `reprex` is an interactive workflow package, don't use this with `dplyr`, `ggplot2`, etc.

When ever you want to ask people about your code, or ask them to debug it


 1. Copy your code to clipboard
 2. Simply run in your console

```{r, eval = FALSE}
reprex()


reprex(advertise = TRUE, # To suppress ads
       si = TRUE, # to include session info
       style = TRUE, # restyle your code
       tidyverse_quiet = TRUE # suppress tidyverse message
       )
```


Or instead of doing it manually, you can select your code, and under "Addins". select "Render reprex"

