---
title: Connect WRDS in R
author: Mike Nguyen
date: '2020-12-29'
slug: connect-wrds-in-r
categories:
  - Finance
  - Data
  - marketing
  - R
tags:
  - Academic
  - data
  - data manipulation
  - finance
subtitle: ''
summary: ''
authors: []
lastmod: '2020-12-29T17:10:53-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

Connect from R to Wharton Research Data Services

to set up connection from R to WRDS ([here](https://wrds-www.wharton.upenn.edu/pages/support/programming-wrds/programming-r/r-from-your-computer/))

```{r, message=FALSE}
library(RPostgres)
library(dplyr)


# I've set up wrds connection before hand. 
# Please use your username and password here.
wrds <- dbConnect(
  Postgres(),
  host = 'wrds-pgdata.wharton.upenn.edu',
  port = 9737,
  dbname = 'wrds',
  sslmode = 'require',
  user = Sys.getenv("wrds_user"),
  pass = Sys.getenv("wrds_pass")
)
knitr::opts_chunk$set(message = FALSE)
```

Example

Check variables (column headers) in COMP ANNUAL FUNDAMENTAL

```{r}
res <- dbSendQuery(
  wrds,
  "select column_name
                   from information_schema.columns
                   where table_schema='compa'
                   and table_name='funda'
                   order by column_name"
)
# data <- dbFetch(res, n=-1) 
data <- dbFetch(res, n = 10)
dbClearResult(res) # closes the connection
head(data)
```

## Overview

All data libraries available at WRDS

```{r}
res <- dbSendQuery(wrds, "select distinct table_schema
                   from information_schema.tables
                   order by table_schema")
all_lib <- dbFetch(res, n=-1)
dbClearResult(res)
head(all_lib)
```

datasets within a given library

```{r}
res <- dbSendQuery(wrds, "select distinct table_name
                   from information_schema.columns
                   where table_schema='comp_na_daily_all'
                   order by table_name")
df_in_lib <- dbFetch(res, n=-1)
dbClearResult(res)
head(df_in_lib)
```

variables within a given dataset

```{r}
res <- dbSendQuery(wrds, "select column_name
                   from information_schema.columns
                   where table_schema='comp_na_daily_all'
                   and table_name='funda'
                   order by column_name")
variables <- dbFetch(res, n=-1)
dbClearResult(res)
head(variables)
```

## Specific Example

Get advertising and R&D data

```{r}
# from comp_na_daily_all.funda
res <- dbSendQuery(wrds, "select gvkey, fyear, xad, xrd
                   from comp_na_daily_all.funda
                   where fyear between 2000 and 2010 
                     and xad is not null 
                     and xrd is not null")

# res <- dbSendQuery(wrds, "select gvkey, fyear
#                    from comp_na_daily_all.funda
#                    where fyear between 2000 and 2020")

data <- dbFetch(res, n=10) |> 
# data <- dbFetch(res, n=-1) |> 
    unique()
dbClearResult(res)
head(data)
```

Get industry data

In WRDS, the industry classifications visible on the web version differ from those in the code version. This discrepancy arises because the web dataset undergoes (behind the scene) a merging process where different variables are combined to produce the final table. Consequently, in the library dataset `comp_na_daily_all.funda`, key classifications such as `gind`, `gsubind`, `naics`, and `sic` are absent. Instead, it includes naics historical and sic historical, which often contain incomplete data. To obtain a complete view, users must merge this dataset with the [industry dataset](https://wrds-www.wharton.upenn.edu/data-dictionary/comp_na_daily_all/names/).

```{r}
# get industry data
res <- dbSendQuery(wrds,
                   "select gvkey, gind, gsubind, naics, sic
                   from comp_na_daily_all.names")
ind <- dbFetch(res, n = -1)
dbClearResult(res)
head(ind)
```

```{r}
df <- data |> 
    left_join(ind)
head(df)
```
