---
title: "If You Roll Out Market by Market, Your Lift Estimate Is Probably Too Low"
author: Mike Nguyen
date: '2026-08-19'
slug: staggered-rollouts-break-twfe
categories:
  - R
  - Python
  - Causal Inference
tags:
  - causal inference
  - experimentation
  - marketing analytics
  - difference-in-differences
subtitle: 'Staggered geo rollouts break the regression almost everyone reaches for first.'
summary: 'The two-way fixed effects regression is the default for staggered geo rollouts. On simulated data where the true lift is known, it understates the effect by 41%. Here is why, and the estimator that fixes it, in R and Python.'
authors: []
lastmod: '2026-08-19T09:00:00-07:00'
featured: no
draft: false
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```

Geo experiments are having a moment. With a large share of individual-level conversion
signal now gone, measurement has moved back up to the market level: split the country into
regions, turn the campaign on in some of them, compare. It survives signal loss because it
never needed to track a person in the first place.

But almost nobody turns a campaign on everywhere at once. Budget, inventory, and legal
approvals mean markets go live in waves: ten DMAs in March, ten more in May, the rest in
July. That is a *staggered rollout*, and it quietly breaks the regression that most teams
reach for first.

Here is the problem on data where I know the right answer, because I generated it.

## A rollout where the truth is known

Three hundred markets, twenty weeks. Markets go live in three waves (weeks 8, 12, and 16),
and a quarter of them never launch at all. The campaign builds. It lifts sales by 0.30 in
its first week and grows with exposure, which is what a real campaign does as awareness
accumulates.

```{r simulate}
library(data.table)
library(fixest)
library(did)

set.seed(2026)
n_geo <- 300
n_per <- 20

geo <- data.table(
  geo_id = 1:n_geo,
  g      = sample(c(8, 12, 16, Inf), n_geo, replace = TRUE, prob = rep(.25, 4)),
  geo_fe = rnorm(n_geo, 0, 1)
)

panel <- CJ(geo_id = 1:n_geo, week = 1:n_per)
panel <- merge(panel, geo, by = "geo_id")

panel[, exposure := fifelse(is.finite(g) & week >= g, week - g, NA_real_)]
panel[, tau      := fifelse(is.na(exposure), 0, 0.30 * (exposure + 1))]
panel[, treated  := as.integer(!is.na(exposure))]
panel[, sales    := 10 + geo_fe + 0.05 * week + tau + rnorm(.N, 0, 0.5)]

true_att <- panel[treated == 1, mean(tau)]
```

Averaged over every treated market-week, the true average effect is
**`r sprintf("%.2f", true_att)`**. That is the number any honest estimator should return.

## The default answer is far too low

The standard move is a two-way fixed effects regression with market effects, week effects,
and one treatment dummy:

```{r twfe}
m_twfe <- feols(sales ~ treated | geo_id + week, data = panel)
twfe   <- coef(m_twfe)[["treated"]]
twfe
```

```{r twfe-gap, echo=FALSE}
cat(sprintf("True ATT : %.3f\nTWFE     : %.3f\nBias     : %.1f%%\n",
            true_att, twfe, 100 * (twfe - true_att) / true_att))
```

The regression is fine. The standard error is small. Nothing in the output looks wrong. And
the answer is off by roughly **`r sprintf("%.0f%%", abs(100 * (twfe - true_att) / true_att))`**
in the direction that costs you budget. It makes the campaign look weaker than it is.

## Why it happens

Two-way fixed effects doesn't compute one comparison. It computes a weighted average of
*every* two-group, two-period comparison available in the panel, and some of those
comparisons use **already-treated markets as the control group**.

That is the flaw. When the wave-2 markets go live in week 12, the regression happily uses
the wave-1 markets, treated since week 8 and by now four weeks into a *growing* effect, as a
comparison. Their outcome is still rising because of the campaign. Subtracting that rise
treats real campaign lift as if it were the baseline trend, and the estimate gets dragged
down. [Goodman-Bacon (2021)](https://doi.org/10.1016/j.jeconom.2021.03.014) decomposed
exactly this. Those bad comparisons can even enter with negative weight, which is how a
genuinely positive campaign can produce a negative coefficient.

The condition that triggers it is not exotic. You need only two things: **staggered timing**
and **effects that change with exposure**. Every campaign rollout I have seen has both.

## The fix

[Callaway and Sant'Anna (2021)](https://doi.org/10.1016/j.jeconom.2020.12.001) estimate the
effect separately for each cohort at each period, using only clean controls, meaning markets
that are not yet live, and then aggregate. In R that is the `did` package:

```{r cs}
panel[, gvar := fifelse(is.finite(g), g, 0)]   # 0 codes "never treated"

cs <- att_gt(
  yname = "sales", tname = "week", idname = "geo_id", gname = "gvar",
  data = as.data.frame(panel),
  control_group = "notyettreated",
  bstrap = TRUE, cband = FALSE
)

simple <- aggte(cs, type = "simple", na.rm = TRUE)
```

```{r cs-out, echo=FALSE}
cat(sprintf("True ATT      : %.3f\nTWFE          : %.3f  (bias %+.1f%%)\nCallaway-SA   : %.3f  (bias %+.1f%%)\n",
            true_att,
            twfe, 100 * (twfe - true_att) / true_att,
            simple$overall.att, 100 * (simple$overall.att - true_att) / true_att))
```

That recovers the truth to within sampling error. The small remaining gap is weighting.
`aggte(type = "simple")` weights group-time cells by their size, which is not identical to
averaging over treated market-weeks. It is noise rather than bias, about
`r sprintf("%.1f", abs(simple$overall.att - true_att) / simple$overall.se)` standard errors,
against a TWFE gap of
`r sprintf("%.0f", abs(twfe - true_att) / se(m_twfe)[["treated"]])`.

## Read the curve, not the scalar

The single number was never the interesting part. Because the effect grows, what you
actually want is the shape:

```{r event-study, echo=FALSE}
library(ggplot2)
dyn <- aggte(cs, type = "dynamic", na.rm = TRUE)
es <- data.frame(e = dyn$egt, att = dyn$att.egt, se = dyn$se.egt)
es <- es[es$e >= -5 & es$e <= 6, ]
es$truth <- ifelse(es$e < 0, 0, 0.30 * (es$e + 1))

ggplot(es, aes(e, att)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_vline(xintercept = -0.5, linetype = "dashed", colour = "grey60") +
  geom_line(aes(y = truth), colour = "#C0392B", linewidth = .8, linetype = "22") +
  geom_pointrange(aes(ymin = att - 1.96 * se, ymax = att + 1.96 * se),
                  colour = "#1F4E79", size = .35) +
  geom_hline(yintercept = twfe, colour = "#7F8C8D", linewidth = .6) +
  annotate("text", x = 6, y = twfe - .18, label = "single TWFE estimate",
           colour = "#7F8C8D", size = 3, hjust = 1) +
  annotate("text", x = 6, y = 2.35, label = "true effect",
           colour = "#C0392B", size = 3, hjust = 1) +
  labs(x = "Weeks since market went live", y = "Effect on sales",
       title = "The campaign builds. One coefficient cannot say that.",
       subtitle = "Callaway-Sant'Anna event study (points) vs. truth (dashed)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

Pre-launch estimates sit flat at zero, which is the design check. After launch the effect
climbs steadily, and the flat grey line shows what the single TWFE coefficient claims
instead. A campaign that pays back over six weeks and one that pays back instantly can
produce the same scalar. Only one of them justifies a longer flight.

## The same trap in Python

Nothing here is an R problem. It is a data problem, and the identical bias appears in
Python. For a balanced panel you don't even need a package, because the two-way within
estimator is just OLS on the double-demeaned series.

```python
import numpy as np, pandas as pd

rng = np.random.default_rng(2026)
n_geo, n_per = 300, 20

g      = rng.choice([8, 12, 16, np.inf], size=n_geo, p=[.25] * 4)
geo_fe = rng.normal(0, 1, n_geo)

geo_id = np.repeat(np.arange(n_geo), n_per)
week   = np.tile(np.arange(1, n_per + 1), n_geo)

df = pd.DataFrame({"geo_id": geo_id, "week": week,
                   "g": g[geo_id], "geo_fe": geo_fe[geo_id]})

df["exposure"] = np.where(df.week >= df.g, df.week - df.g, np.nan)
df["tau"]      = np.where(df.exposure.notna(), 0.30 * (df.exposure + 1), 0.0)
df["treated"]  = df.exposure.notna().astype(int)
df["sales"]    = (10 + df.geo_fe + 0.05 * df.week + df.tau
                  + rng.normal(0, 0.5, len(df)))

def twoway_demean(col):
    x = df[col] - df[col].mean()
    x = x - x.groupby(df.geo_id).transform("mean")
    x = x - x.groupby(df.week).transform("mean")
    return x

y, d = twoway_demean("sales"), twoway_demean("treated")
twfe = (d @ y) / (d @ d)

print(f"True ATT : {df.loc[df.treated == 1, 'tau'].mean():.3f}")
print(f"TWFE     : {twfe:.3f}")
```

Same understatement, different random draws. R and NumPy do not share a generator, so the
digits will not match, only the conclusion. For the Callaway-Sant'Anna estimator itself,
Python has [`differences`](https://github.com/bernardodionisi/differences) and
[`diff-diff`](https://pypi.org/project/diff-diff/). The latter also covers synthetic DiD and
Honest DiD sensitivity analysis.

## What to do on Monday

If you are measuring a staggered rollout:

1. **Check whether your timing is actually staggered.** If every market launched the same
   week, plain DiD is fine and none of this applies.
2. **Assume your effect is dynamic.** Awareness builds, and so does the bias. Treating the
   effect as one constant number is the assumption that breaks TWFE, not the staggering
   itself.
3. **Never let an already-treated market serve as a control.** This is the whole bug. Use
   not-yet-treated or never-treated markets.
4. **Report the event-study curve, not one coefficient.** The shape is the decision-relevant
   output. It tells you the payback period.
5. **Treat flat pre-trends as a check you passed, not a proof.** Parallel trends is an
   assumption about an unobservable counterfactual. If the decision is expensive, run a
   sensitivity analysis rather than pointing at the pre-period.

The uncomfortable version of this: if you have been sizing incrementality off staggered
rollouts with a TWFE regression, your campaigns may have looked systematically weaker than
they are, and the budget decisions that followed inherited that error.

---

*Estimator details, the Goodman-Bacon decomposition, and the assumptions behind each of these
designs are covered in [Experimental Design](https://link.springer.com/book/9783032018380),
Volume 4 of A Guide on Data Analysis.*
