---
title: "Power Analysis for Interaction Effects in R: Your 80% Powered Experiment Is 15% Powered for the Moderator"
author: Mike Nguyen
date: '2026-08-28'
slug: power-analysis-for-interaction-effects-in-r
categories:
  - R
  - Experimental Design
tags:
  - power analysis
  - interaction effects
  - moderation
  - experimental design
  - type m error
subtitle: 'The sample size that detects your treatment effect is nowhere near the sample size that detects who it works for.'
summary: 'Power analysis for interaction effects in R, done by simulation. A study sized for 80% power on the main effect delivers 15% power on the interaction, and the interactions that do clear significance overstate the truth by 2.6x. The multiplier you need is four times the squared ratio of main effect to interaction.'
authors: []
lastmod: '2026-08-28T11:30:00-07:00'
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.2, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```

Every experiment ends with someone asking who the effect was strongest for. The power analysis
that sized it answered a different question, so the honest reply is usually that the data cannot
say. Power analysis for interaction effects is the calculation almost nobody runs,
and skipping it does not merely cost you a finding. It manufactures fake ones.

Here is the whole problem in two numbers. Size a study for 80% power on the treatment effect,
then ask that same study whether the effect differs between two segments.

```{r headline, echo=FALSE}
library(pwr)
d_main <- 0.30                # average treatment effect, in SD units
tau_A  <- 0.20                # effect in segment A
tau_B  <- 0.40                # effect in segment B
inter  <- tau_B - tau_A       # the interaction

# Exact power from the non-central t, which is the test lm() actually runs
pw <- function(eff, se, df) {
  tc <- qt(0.975, df)
  pt(tc, df, ncp = eff / se, lower.tail = FALSE) + pt(-tc, df, ncp = eff / se)
}
n_arm <- ceiling(pwr.t.test(d = d_main, power = 0.80, sig.level = 0.05)$n)
N     <- 2 * n_arm
pow_M <- pw(d_main, 2 / sqrt(N), N - 2)   # SE of a main effect,  derived below
pow_I <- pw(inter,  4 / sqrt(N), N - 4)   # SE of an interaction, derived below

data.frame(
  test  = c("main effect: does the treatment work",
            "interaction: does it work differently by segment"),
  truth = c(d_main, inter),
  power = round(c(pow_M, pow_I), 3)
)
```

Same participants, same experiment. One test is a bet you win four times out of five. The other
is a bet you lose more than five times out of six.

## The study you actually designed

A standardized effect of 0.30 is an ordinary ambition for a marketing or behavioral
intervention. The `pwr` package sizes it in one line.

```{r design}
library(pwr)

d_main <- 0.30
pwr.t.test(d = d_main, power = 0.80, sig.level = 0.05)
```

```{r nsize}
n_arm <- ceiling(pwr.t.test(d = d_main, power = 0.80, sig.level = 0.05)$n)
N     <- 2 * n_arm
c(n_per_arm = n_arm, N_total = N)
```

So `r N` people. Now the interaction. Suppose the treatment genuinely works better for segment B
than for segment A, with effects of 0.40 and 0.20. Their average is the 0.30 the study was built
around, so the design is not wrong about the main effect. The interaction is the difference.

```{r truth}
tau_A <- 0.20
tau_B <- 0.40
inter <- tau_B - tau_A
c(main_effect = (tau_A + tau_B) / 2, interaction = inter)
```

## What that sample buys for the interaction

Rather than trusting a formula, run the study twenty thousand times and count.

```{r mc}
set.seed(20260828)

sim_once <- function(N, tau_A, tau_B) {
  seg <- rep(0:1, each = N / 2)             # half the sample in each segment
  trt <- rep(rep(0:1, each = N / 4), 2)     # randomized within segment
  y   <- ifelse(seg == 1, tau_B, tau_A) * trt + rnorm(N)
  summary(lm(y ~ trt * seg))$coefficients["trt:seg", c(1, 2, 4)]
}

mc <- as.data.frame(t(replicate(20000, sim_once(N, tau_A, tau_B))))
names(mc) <- c("est", "se", "p")
sig <- mc$p < 0.05

round(c(power = mean(sig), mean_SE = mean(mc$se)), 4)
```

`r round(100 * mean(sig))`%, and the simulation agrees with the exact non-central t answer of
`r round(100 * pow_I, 1)`% to within Monte Carlo error. The study that finds the treatment effect
four times out of five finds the moderator less than one time in six.

## Why the standard error doubles

The arithmetic is short enough to do by hand, and doing it once retires the simulation.

Write the four cell means as `m[segment, arm]`. The main effect is a difference of averages.

    A = (m[A,1] + m[B,1])/2 - (m[A,0] + m[B,0])/2

Each cell mean carries weight 1/2, so with `n` per cell the variance is
`4 * (1/2)^2 * sigma^2/n`, which is `sigma^2/n`. The interaction is a difference of differences.

    I = (m[B,1] - m[B,0]) - (m[A,1] - m[A,0])

Now every cell mean carries weight 1, so the variance is `4 * sigma^2/n`, four times larger.

**The standard error of an interaction is exactly twice the standard error of the main effect in
the same experiment.** That factor of two is structural. No amount of clean design removes it,
because the interaction is assembled from four noisy quantities instead of two.

```{r se-check}
c(analytic_SE = 4 / sqrt(N), simulated_SE = mean(mc$se))
```

Two things now work against you. The standard error is doubled, and the interaction is usually
smaller than the main effect it modifies. Setting the two z-statistics equal gives the
multiplier.

    N_interaction / N_main = 4 * (main effect / interaction)^2

```{r multiplier}
mult <- function(main, inter) 4 * (main / inter)^2

z80 <- qnorm(0.975) + qnorm(0.80)
N_I <- (4 * z80 / inter)^2                 # N for 80% power on the interaction

c(exact_N_needed = ceiling(N_I),
  exact_ratio    = round(N_I / N, 2),
  formula_ratio  = mult(d_main, inter))
```

Nine times the sample for the same power, on the same question everyone asks at the end of the
readout. The folklore number is 16, and this shows where 16 comes from. It is
the special case where the interaction happens to be exactly half the main effect.

```{r folklore}
data.frame(
  interaction_relative_to_main = c("equal", "three quarters", "half", "one third", "one quarter"),
  multiplier = mult(1, c(1, 0.75, 0.5, 1/3, 0.25))
)
```

A moderator worth a quarter of the main effect costs 64 times the sample. That is not a study you
resize. That is a study you redesign.

```{r curve, echo=FALSE, fig.cap="Power for the main effect and for the interaction, on the same participants."}
library(ggplot2)

Ns <- seq(100, 4000, by = 20)
plot_df <- rbind(
  data.frame(N = Ns, test = "Main effect (0.30)",  power = pw(d_main, 2 / sqrt(Ns), Ns - 2)),
  data.frame(N = Ns, test = "Interaction (0.20)",  power = pw(inter,  4 / sqrt(Ns), Ns - 4))
)
pts <- data.frame(N = c(N, N),
                  test = c("Main effect (0.30)", "Interaction (0.20)"),
                  power = c(pow_M, pow_I))

ggplot(plot_df, aes(N, power, colour = test)) +
  geom_hline(yintercept = 0.8, linetype = 2, colour = "grey55") +
  geom_line(linewidth = 1) +
  geom_point(data = pts, size = 2.8) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("Main effect (0.30)" = "#2c6fbb",
                                 "Interaction (0.20)" = "#c1502e")) +
  labs(x = "Total sample size", y = "Power", colour = NULL,
       title = paste0("At N = ", N, ", the same experiment is ",
                      round(100 * pow_M), "% and ", round(100 * pow_I), "% powered")) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.title = element_text(size = 11.5))
```

## The part that turns weakness into damage

Low power on its own is a wasted test. What makes an underpowered interaction actively harmful is
what survives the significance filter. Look only at the runs that reached p < 0.05, the only ones
anyone would write up.

```{r type-m}
round(c(
  truth             = inter,
  mean_estimate_all = mean(abs(mc$est)),
  mean_estimate_sig = mean(abs(mc$est[sig])),
  exaggeration      = mean(abs(mc$est[sig])) / inter,
  wrong_sign_rate   = mean(mc$est[sig] < 0)
), 3)
```

The true interaction is 0.20. Conditional on clearing significance, the average estimate is
`r round(mean(abs(mc$est[sig])), 2)`, an exaggeration of
**`r round(mean(abs(mc$est[sig])) / inter, 1)`x**. This is Type M error, the magnitude error, and
it is a mathematical consequence of filtering noisy estimates on significance rather than a
symptom of anything sloppy in the analysis.

That is what a published subgroup effect looks like when it comes out of a study sized for the
main effect. Not absent. Inflated by enough to move a budget. The replication then returns a
third of the size and gets called a failure to replicate, when the original was simply the tail
of a distribution centered on something much smaller.

About `r round(100 * mean(mc$est[sig] < 0), 1)`% of the significant results also point the wrong
way, which is small here only because the true interaction is not near zero.

## Do not throw away a continuous moderator

The cheapest recovery is to stop dichotomizing. If the moderator is continuous (age, prior spend,
engagement score) and you median-split it, you are discarding information for nothing.

```{r dichotomize}
set.seed(11)
zs <- replicate(3000, {
  n <- 1500
  Tr <- rbinom(n, 1, 0.5)
  M  <- rnorm(n)
  y  <- 0.30 * Tr + 0.20 * Tr * M + rnorm(n)
  D  <- as.integer(M > 0)                    # the median split
  c(summary(lm(y ~ Tr * M))$coefficients["Tr:M", 3],
    summary(lm(y ~ Tr * D))$coefficients["Tr:D", 3])
})

eff <- (mean(zs[2, ]) / mean(zs[1, ]))^2
round(c(relative_efficiency   = eff,
        theoretical_2_over_pi = 2 / pi,
        extra_sample_needed   = 1 / eff - 1), 3)
```

The split retains `r round(100 * eff)`% of the information, matching the theoretical `2/pi` for a
normal moderator, so it costs about `r round(100 * (1 / eff - 1))`% more sample to reach the same
power. When you already need nine times the sample, handing another third to a median split is
the last thing you can afford.

## Power the interaction at design time

The general tool is a simulation, because the closed form stops being convenient the moment the
design has clusters, unequal cells, or a non-normal outcome. This function extends to all of
those cases by editing the line that generates `y`.

```{r power-fn}
power_interaction <- function(N, tau_A, tau_B, sigma = 1,
                              p_seg = 0.5, alpha = 0.05, R = 1500) {
  mean(replicate(R, {
    seg <- rbinom(N, 1, p_seg)
    trt <- rbinom(N, 1, 0.5)
    y   <- ifelse(seg == 1, tau_B, tau_A) * trt + rnorm(N, sd = sigma)
    summary(lm(y ~ trt * seg))$coefficients["trt:seg", 4] < alpha
  }))
}

set.seed(7)
sapply(c(352, 1000, 3140), function(n) power_interaction(n, 0.20, 0.40))
```

The third value confirms that roughly 3,140 participants is where 80% power for this interaction
actually lives. It lands just under 0.80 because this version randomizes segment membership
rather than fixing the cell counts, which is the more honest model of real recruitment and costs
a little precision.

## What to do on Monday

1. **Power the interaction, not just the main effect.** If the study exists to answer "who does
   this work for," that is the primary analysis and it sets the sample size.
2. **Use `4 * (main / interaction)^2` as the napkin check** before anyone builds a simulation. It
   tells you in seconds whether the answer is "add 30%" or "not with this budget."
3. **Write down the interaction you expect, in the same units as the main effect.** Most teams
   have never said this number out loud, and saying it is usually what reveals the study is
   infeasible.
4. **Never median-split a continuous moderator.** It costs about 58% more sample for nothing.
5. **When the required N is out of reach, stop testing and start estimating.** Pre-specify one
   interaction, report the estimate with its interval, and say plainly that the study could not
   resolve it. An honest wide interval beats a test that inflates whatever it certifies.
6. **Treat a significant subgroup effect from a main-effect-sized study as an upper bound**
   rather than as an estimate. The filter guarantees the surviving number came from the
   optimistic tail.
7. **Cut the interactions you look at down to the ones you would have defended in advance.**
   Every extra subgroup is another draw from that same filter.

The uncomfortable conclusion is that a large share of subgroup findings, in journals and in
dashboards alike, come from studies sized for something else. The effect being real does not
rescue the estimate, because the filter operates on the estimate rather than on the truth.

---

*Interaction terms, how to interpret them once they are estimated, and what changes when the
moderator is continuous are covered in the
[moderation chapter](https://bookdown.org/mike/data_analysis/moderation.html)
of A Guide on Data Analysis.*
