---
title: "Tobin's q"
author: "Mike Nguyen"
date: '2020-12-19'
slug: tobin-s-q
categories:
- Finance
- marketing
tags: finance
subtitle: ''
summary: ''
authors: []
lastmod: '2020-12-19T19:54:29-06:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
bibliography: references.bib
---

# Tobin's Q

According to Kerin and Sethuraman [@Kerin_1998, p. 261], **Tobins q** is the "ratio of the market value of the firm to the replacement cost of its tangible assets, property, plant, equipment, inventory, cash, and investments in stocks and bonds".  

**Market-to-book ratio** is the equity capitalization of a firm divided by book equity. Equivalently, (share price multiplied by number of shares) over (common stock equity, capital surplus, and retained earnings).  


**Market-to-book ratio** and **Tobin's q** have been shown to be equivalent measures [@Chung_1994] 

Original Tobin's q calculation [@Lindenberg_1981]  

$$
q = \frac{PREFST + VCOMS + LTDEBT + STDEBT - ADJ}{TOTASST - BKCAP + NETCAP}
$$

where  

 * PREFST is the liquidating value of a firm's preferred stock.  
 * VCOMS is the price of the firm's common stock multiplied by the number of shares outstanding at year end (December 31st).  
 * LTDEBT is the long-term debt adjusted for its age structure 
 * STDEBT is the book value of the firm's current liabilities  
 * ADJ is the net short-term assets  
 * TOTASST is the book value of the firm's total assets  
 * BKCAP is the book value of the firm's net capital stock  
 * NETCAP is information-adjusted net capital stock.  


However, [@Chung_1994] have proposed a simple reliable, and tractable approximation of Tobin's q:  

$$
q = \frac{MVE + PS + DEBT}{TA}
$$

where  

 * MVE = share price x number of common stock share outstanding 
 * PS = liquidating value of outstanding preferred stock  
 * DEBT = short-term liability - short-term assets + book value of long-term debt  
 * TA is the book value of the total assets.  

[@Chung_1994] method assumes that the replacement values of a firm's plant, equipment, and inventories equal to their book values. 

## Application

Retrieve data from WRDS

```{r, echo=FALSE, message=FALSE}
# to set up connection from R to WRDS (https://wrds-www.wharton.upenn.edu/pages/support/programming-wrds/programming-r/r-from-your-computer/)
library(RPostgres)
library(dplyr)
wrds <- dbConnect(Postgres(),
                  host='wrds-pgdata.wharton.upenn.edu',
                  port=9737,
                  dbname='wrds',
                  sslmode='require',
                  user='nghia130')
knitr::opts_chunk$set(message = FALSE)
```



```{r}
# to set up connection from R to WRDS (https://wrds-www.wharton.upenn.edu/pages/support/programming-wrds/programming-r/r-from-your-computer/)
library(RPostgres)
library(dplyr)

# I've set up wrds connection before hand. Please use your username and password here.

# wrds <- dbConnect(Postgres(),
#                   host='wrds-pgdata.wharton.upenn.edu',
#                   port=9737,
#                   dbname='wrds',
#                   sslmode='require',
#                   user='')

# Check variables (column headers) in COMP ANNUAL FUNDAMENTAL
#uses the already-established wrds connection to prepare the SQL query string and save the query as the result res.
# check avaiable databases: https://wrds-web.wharton.upenn.edu/wrds/tools/variable.cfm?vendor_id=7
res <- dbSendQuery(wrds, "select column_name
                   from information_schema.columns
                   where table_schema='compa'
                   and table_name='funda'
                   order by column_name")
data <- dbFetch(res, n=-1) # fetches the data that results from running the query res against wrds and stores it as data
dbClearResult(res) # closes the connection
head(data)

# select everything
res <- dbSendQuery(wrds, "select * from compa.funda")

# from compa.funda

# only select the following variables
res <- dbSendQuery(wrds, "select gvkey, datadate, fyear, indfmt, consol, popsrc, datafmt, tic, cusip, conm, curcd, fyr, act, at, bkvlps, ceq, ch, che, dltt, dlc, emp, np, exchg, cik, costat, naicsh,mkvalt from compa.funda") #check variables from (https://wrds-web.wharton.upenn.edu/wrds/ds/comp/funda/index.cfm?navId=80)
data1 <- dbFetch(res, n=-1)
dbClearResult(res)

data = data1 %>%
    distinct(gvkey,datadate,fyear,tic,conm,.keep_all = T)
```


Calculate Tobin's Q

```{r}
tobin_q_dt = data %>%
    mutate(mkvalt = coalesce(mkvalt,0),
           dltt = coalesce(dltt,0),
           at = coalesce(at,0),
           dlc = coalesce(dlc,0),
           act = coalesce(act,0)) %>%
    mutate(tobin_q = (mkvalt + ifelse((dlc - act) >=0,as.numeric(dlc-act),0) + dltt)/at ) %>% #follow Chung (1994) (sum of stocks (preferred + common) + debt(short-term liabilities - short-term assets + long-term debt))/(total assets) # note: take only excess of short-term liabilities over short-term assets to be included in debt. 
    select(tobin_q,gvkey,datadate,fyear,conm)
head(tobin_q_dt)
```

