---
title: "doParallel"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
categories: R
tags:
- R Markdown
- HPC
---

# Packages

```{r}
#install.packages("doParallel")
library("doParallel")
```

Check how long a loop takes

```{r}
start <- proc.time()
for (i in 1:10){
    sqrt(i)
}
base_loop=(proc.time()-start)
```

Check number of cores
```{r}
library("doParallel")
detectCores()
```

```{r}
getDoParWorkers() #current number of worker
registerDoSEQ() #switch back to original
getDoParWorkers()
```

Let R know how many cores you want to use 
```{r}
registerDoParallel(6)
```

**Note**: You shouldn't use all the cores since your computer might want some background programs to run.

We can also create cluster of workers

```{r}
cluster = makeCluster(2)
registerDoParallel(cluster)
system.time(foreach(i=1:100) %dopar% sum((1:i)))
stopCluster(cluster)
```



Compare different loops

`%do%` 

```{r}
start <- proc.time()
foreach(i=1:10) %do% {
    sqrt(i)
}
do_loop=(proc.time()-start)

```


`%dopar%` loop

```{r}
cl <- makeCluster(2)
registerDoParallel(cl)
start <- proc.time()
foreach(i = 1:10, .combine = 'c') %dopar% {
  sqrt(i)
}
dopar_loop=(proc.time()-start)
stopCluster(cl)
```


```{r}
print(rbind(base_loop,do_loop,dopar_loop)[,1:3])
```

`%do%` does not use more than base core, but `%dopar%` uses the number of cores that we specify in `registerDoParallel`

`doParallel::mclapply` is the equivalent of `lapply`.