---
title: "Unicode Normalization and LLM Token Costs: The Same Korean Word Is 2 Tokens or 36"
author: Mike Nguyen
date: '2026-08-28'
slug: unicode-normalization-and-llm-token-costs
categories:
  - R
  - AI
tags:
  - llm
  - tokenization
  - unicode
  - text analysis
  - reticulate
subtitle: 'Two strings that render identically on screen can differ eighteenfold in what you pay for them.'
summary: 'A Unicode normal form you have probably never checked can multiply your LLM token bill by ten. The same Korean greeting costs 2 tokens composed and 36 tokens decomposed, and one of the eleven translations in the Unicode Consortium own reference corpus ships decomposed. Measured in R with stringi and tiktoken.'
authors: []
lastmod: '2026-08-28T15:00:00-07:00'
featured: no
draft: true
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 7, fig.height = 4.4, dpi = 150,
                      # NOT the default `*_files/figure-html/`. The ignoreFiles
                      # setting in config.toml drops `_files$`, so images written
                      # there never reach the published site.
                      fig.path = "figs/")
```


You pay LLM providers by the token, so token count is the price of everything you send and
everything you get back. Two strings that are indistinguishable on screen, that compare equal to
a human reader and pass every eyeball test, can differ eighteenfold in what they cost. The
difference is a Unicode normal form nobody on your team has checked.

Here is the Korean word for hello, tokenized twice.

```{r hello}
library(reticulate)
library(stringi)

enc <- import("tiktoken")$get_encoding("o200k_base")   # GPT-4o and later
ntok <- function(s) length(enc$encode(s))

hello <- "안녕하세요"

c(composed_NFC   = ntok(stri_trans_nfc(hello)),
  decomposed_NFD = ntok(stri_trans_nfd(hello)))
```

```{r hello-ratio, include=FALSE}
hello_ratio <- ntok(stri_trans_nfd(hello)) / ntok(stri_trans_nfc(hello))
```

Same word. Same rendering in your browser, your terminal, and your database client. One costs
`r round(hello_ratio)` times more than the other.

## What the two forms are

Unicode lets you spell an accented character two ways. **NFC** composes it into a single code
point. **NFD** decomposes it into a base letter followed by combining marks. Korean goes
further: a Hangul syllable like 안 is one code point composed, and three separate jamo
decomposed.

The two forms are *canonically equivalent*, which is the Unicode standard's way of saying they
mean the same thing and must display the same way.

```{r equivalence}
nfc <- stri_trans_nfc(hello)
nfd <- stri_trans_nfd(hello)

identical(stri_trans_nfc(nfd), nfc)          # same text once you re-compose
identical(charToRaw(nfc), charToRaw(nfd))    # same bytes on the wire
c(codepoints_nfc = stri_length(nfc), codepoints_nfd = stri_length(nfd))
```

Equivalent to a reader, different to a tokenizer. String equality in most languages compares
bytes, so these two also fail `==` in the databases and caches sitting in front of your model.

## How big is it across real text

Article 1 of the Universal Declaration of Human Rights, in eleven languages, taken verbatim from
the Unicode Consortium's own reference corpus. Each one tokenized composed and decomposed.

The corpus sits next to this post as a tab separated file, written straight from the upstream XML
and never opened in an editor. That detail matters more than it sounds: editors, clipboards, and
some version control settings silently recompose text, which would quietly erase the exact thing
being measured here.

```{r corpus}
udhr <- read.delim("udhr_article1.tsv", encoding = "UTF-8",
                   quote = "", stringsAsFactors = FALSE)
udhr <- setNames(udhr$text, udhr$language)
length(udhr)
```

```{r table}
tok_both <- function(s) {
  c(nfc = ntok(stri_trans_nfc(s)), nfd = ntok(stri_trans_nfd(s)))
}

res <- t(sapply(udhr, tok_both))
res <- data.frame(
  language = rownames(res),
  nfc      = res[, "nfc"],
  nfd      = res[, "nfd"],
  penalty  = round(res[, "nfd"] / res[, "nfc"], 2),
  vs_english = round(res[, "nfc"] / res["English", "nfc"], 2),
  row.names = NULL
)
res[order(-res$penalty), ]
```

Korean pays `r res$penalty[res$language == "Korean"]` times more decomposed than composed.
Vietnamese pays `r res$penalty[res$language == "Vietnamese"]` times more.
Several languages pay nothing at all, and the reason they are safe is that they have no composed
forms to decompose: Chinese, Japanese kana, Russian Cyrillic, and Devanagari are already written
as separate code points.

Now look at the last column, because it is the one that calibrates the whole post. The gap
between languages, once everything is normalized properly, tops out near a factor of two. The
gap between normal forms *within a single language* reaches ten. **Normalization is the larger
lever, and unlike the language your customers speak, it is one you control.**

```{r plot, echo=FALSE, fig.cap="Token cost of the same sentence, composed against decomposed. Where only one point is visible the two forms are identical, because that script has no composed characters to pull apart."}
library(ggplot2)

d <- res[order(res$penalty), ]
d$language <- factor(d$language, levels = d$language)

ggplot(d, aes(y = language)) +
  geom_segment(aes(x = nfc, xend = nfd, yend = language),
               colour = "grey70", linewidth = 1.1) +
  # NFD first, NFC on top, so the languages that pay nothing read as green
  geom_point(aes(x = nfd, colour = "Decomposed (NFD)"), size = 2.9) +
  geom_point(aes(x = nfc, colour = "Composed (NFC)"), size = 2.9) +
  geom_text(data = subset(d, penalty > 1.3),
            aes(x = nfd, label = paste0(penalty, "x")),
            hjust = -0.35, size = 3.3, colour = "#b5451b") +
  scale_x_continuous(limits = c(0, 560), expand = expansion(mult = c(0.01, 0.08))) +
  scale_colour_manual(values = c("Composed (NFC)" = "#2f6f4f",
                                 "Decomposed (NFD)" = "#b5451b")) +
  labs(x = "Tokens for the same sentence", y = NULL, colour = NULL,
       title = "The same text, two Unicode normal forms") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        plot.title = element_text(size = 11.5))
```

## Why decomposing is so expensive

A byte pair encoding vocabulary is built from text as it usually appears, which is composed. The
composed syllable 안 earns its own entry. An isolated jamo almost never appears in training text,
so it earns nothing, and the tokenizer falls back to encoding it as raw UTF-8 bytes.

```{r mechanism}
one <- "안"
rbind(NFC = c(bytes = length(charToRaw(stri_trans_nfc(one))),
              tokens = ntok(stri_trans_nfc(one))),
      NFD = c(bytes = length(charToRaw(stri_trans_nfd(one))),
              tokens = ntok(stri_trans_nfd(one))))
```

Three jamo, three bytes each, one token per byte. That is the ten times, and it is not a quirk of
Korean so much as the general behaviour of any tokenizer meeting characters outside its
vocabulary.

## This is not a hypothetical

The tempting response is that nobody actually stores decomposed text. Check the corpus this post
is built from, which is maintained by the Unicode Consortium itself.

```{r wild}
already_nfc <- sapply(udhr, function(s) identical(s, stri_trans_nfc(s)))
names(already_nfc)[!already_nfc]
```

`r sum(!already_nfc)` of the `r length(udhr)`
translations arrives decomposed, in the reference corpus for Unicode text, published by the
organization that defines Unicode. Decomposed text also comes from macOS, whose filesystem
normalized filenames to NFD for years, from several input methods for Korean and Vietnamese, and
from any pipeline that decomposes for accent-insensitive search and forgets to recompose before
storing.

You will not see it. That is the whole problem. It renders correctly at every step.

## The fix is one line

```{r fix}
raw   <- udhr[["Vietnamese"]]          # as it arrived
clean <- stri_trans_nfc(raw)           # the whole fix

c(tokens_before = ntok(raw),
  tokens_after  = ntok(clean),
  percent_saved = round(100 * (1 - ntok(clean) / ntok(raw))))
```

That is a `r round(100 * (1 - ntok(stri_trans_nfc(udhr[["Vietnamese"]])) / ntok(udhr[["Vietnamese"]])))`% reduction on real upstream text, from one function call, with no change to
a single character anyone can see.

In Python it is `unicodedata.normalize("NFC", s)`. In R it is `stringi::stri_trans_nfc()`. Put it
at the boundary where text enters your system, not next to the API call, so that everything
downstream (your cache keys, your deduplication, your embeddings) agrees on one spelling.

## What to do on Monday

1. **Normalize to NFC on ingest.** One line, at the edge, before anything is stored or hashed.
   Doing it at the API call leaves your cache and your database disagreeing with each other.
2. **Audit what you already have.** Run the `identical(s, stri_trans_nfc(s))` check over a sample
   of stored prompts and documents. It is cheap and it is the only way to find this.
3. **Suspect this first when token counts look wrong for one locale.** A Korean or Vietnamese
   segment costing several times what you modeled is this bug until proven otherwise.
4. **Do not blame the language.** Properly normalized, the spread across languages is roughly
   twofold, which is a real fairness issue but a much smaller number than the one you can fix
   this afternoon.
5. **Normalize before you hash, cache, or deduplicate.** Two canonically equivalent strings are
   different bytes, so a prompt cache will miss and you will pay twice for the same request.
6. **Check the fix survived.** Some libraries re-decompose on write. Assert the invariant in a
   test rather than trusting it.

The reason this survives in production is that every layer looks correct. The text displays
correctly, the database returns it correctly, the model answers correctly. The only thing that
is wrong is the bill.

---

*Tokenization, encoding, and the preprocessing decisions that quietly change what a model sees
are covered in [AI in Action](https://mikenguyen13.github.io/ai_in_action/).*
